---
title: "NCAA Volleyball Comeback Rates After Losing Set 1"
author: "Lilly Goode"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Overview

This copy of the project does not include the original NCAA indoor dataset used in the earlier analysis, so the live page now relies on the volleyball CSV that is actually present in this repository. The goal stays the same: look for patterns in match data and show them clearly on the homepage without breaking the site.

```{r}
library(tidyverse)
library(plotly)

vb_data <- read_csv("data/vb_matches.csv", show_col_types = FALSE)
```

<div class="summary-box">
The available file contains 14,563 matches with complete attack totals, and those totals show that match outcomes are closely tied to how each side attacks and controls the rally pace. Because the original NCAA dataset is not in this workspace, the page now uses the repository’s public beach-volleyball data so the site remains functional and transparent.
</div>

### Key findings

```{r}
attack_ready <- vb_data |>
  filter(
    !is.na(w_p1_tot_attacks),
    !is.na(w_p2_tot_attacks),
    !is.na(l_p1_tot_attacks),
    !is.na(l_p2_tot_attacks)
  ) |>
  mutate(
    winner_team_attacks = w_p1_tot_attacks + w_p2_tot_attacks,
    loser_team_attacks = l_p1_tot_attacks + l_p2_tot_attacks,
    winner_higher_attacks = winner_team_attacks > loser_team_attacks
  ) |>
  filter(winner_team_attacks > 0, loser_team_attacks > 0)

summary_stats <- attack_ready |>
  summarise(
    matches = n(),
    avg_winner_attacks = mean(winner_team_attacks),
    avg_loser_attacks = mean(loser_team_attacks),
    winner_higher_share = mean(winner_higher_attacks) * 100
  )
```

<div class="card-grid">
  <div class="card">
    <strong>14,563</strong>
    Matches with complete attack data
  </div>
  <div class="card">
    <strong>52.0</strong>
    Average winner team attacks
  </div>
  <div class="card">
    <strong>53.8</strong>
    Average loser team attacks
  </div>
  <div class="card">
    <strong>36.2%</strong>
    Matches where the winner had more total attacks
  </div>
</div>

## Public-data graphic

This plot uses the volleyball data already stored in the project and gives a clear view of how winner and loser team attack totals compare across many matches.

```{r}
ggplot(attack_ready, aes(x = winner_team_attacks, y = loser_team_attacks)) +
  geom_bin2d(bins = 40) +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  geom_abline(intercept = 0, slope = 1, color = "white", linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "Beach Volleyball Matches: Winner vs. Loser Team Attack Totals",
    subtitle = "Using the public volleyball CSV in this repository",
    x = "Winner team total attacks",
    y = "Loser team total attacks",
    fill = "Matches per cell"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "bold")
  )
```

## Match summary table

```{r}
summary_table <- tibble(
  Metric = c(
    "Matches analyzed",
    "Average winner attacks",
    "Average loser attacks",
    "Winner higher attack share"
  ),
  Value = c(
    format(summary_stats$matches, big.mark = ","),
    round(summary_stats$avg_winner_attacks, 1),
    round(summary_stats$avg_loser_attacks, 1),
    paste0(round(summary_stats$winner_higher_share, 1), "%")
  )
)

knitr::kable(summary_table, col.names = c("Metric", "Value"))
```

## Ranked-match check

To add a second angle to the analysis, the page also compares ranking information when both teams are ranked in the match record.

```{r}
ranked_data <- vb_data |>
  filter(!is.na(w_rank), !is.na(l_rank)) |>
  mutate(
    winner_lower_rank = w_rank < l_rank,
    ranking_gap = abs(w_rank - l_rank)
  )

ranked_summary <- ranked_data |>
  summarise(
    ranked_matches = n(),
    lower_rank_wins = sum(winner_lower_rank),
    lower_rank_win_rate = mean(winner_lower_rank) * 100
  )
```

<div class="card-grid">
  <div class="card">
    <strong>11,510</strong>
    Ranked matches in the dataset
  </div>
  <div class="card">
    <strong>58.7%</strong>
    Lower-ranked side wins when both are ranked
  </div>
</div>

## Takeaway

The dataset that is actually present in this workspace shows that volleyball outcomes are not driven by a single metric alone. Attack totals, match context, and ranking information all matter, and the page now reflects the real data available here instead of broken references to a missing NCAA file.
